The Impact of Past Behavior Normality on Regret: A Replication Study

### 2024-02-13 outputs/paper/paper.Rmd
update

---
title: "Declining Birth rates in the US: An Analysis of Potential Factors"
subtitle: "Reproduction of 'The Puzzle of Falling US Birth Rates since the Great Recession' (Kearney, Levine & Pardue, 2022)"
author: Heng Ma
thanks: "Code and data are available at: https://github.com/Brian031205/paper_2.git. A replication of various aspects in this paper are available at: ______________    "
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Over the past 10 years, US birth rates have been dropping steadily to an all time low of 55.8 per 1,000 women in 2020. Kearney, Levine and Pardue (2022) analyzed this phenomenon by exploring demographic, economic, social and policy factors to little enlightenment of the cause of the decline. However, they did discover a correlation between the decline and the birth cohorts of mothers, and offer the conjecture of shifting priorities as the reasoning. We replicate the results of this study with respect to the demographic and cohort effects, and discuss alternative interpretations of the data considering the impact of the feminist movement and the evolution of technology and media. The results were consistent with the initial study."
output:
  bookdown::pdf_document2
toc: TRUE
bibliography: references.bib
---

@@

<!-- Background / Broader Context -->
Birth rate is the number of live births per thousand in the population by year [@dict]. Birth rate is important since it is an indicator of population growth and determines the age structure of populations, which has various implications for the economy and society [@CLELAND2008364]. When the birth rate is too high or low, it affects all subgroups of the population in different ways, for example, a large number of new births will be a burden on the adult population to support them, and when this group ages to be elderly they will be a burden on governments to support them [@CLELAND2008364].

There are many possible factors that can affect birth rate. Recessions have been proposed to result in a drop in birth rates, specifically the Great Recession of 2007 [@pew; @focus; @econ]. @becker proposes an economic framework for fertility, where the "demand for children" is driven by the satisfaction that children brings to people weighted against factors such as preferences (e.g. religion, race, age), income, time, expenditure and means to support each child and other considerations (e.g. contraception, marriage). @og base their analysis of the declining US birth rates on these factors.

These analyses however, did not provide sufficient evidence to explain the decline in birth rates, with the authors stating "For any factor to have explained much of that decline, it would have had to change dramatically around the same time." [@og]. @og then went on to find a correlation between the recent birth cohorts of mothers and the substantial decline. In our reproduction of their paper, we build on their 'shifting priorities' explanation of the cohort effects by considering the other notable events that potentially may have influenced the nature of parenting of these cohorts, namely the Third wave of Feminism and the second half of the Digital Revolution.

We replicate the paper by @og with a focus on the following research questions:

* What is the trend in US birth rates over the period 1980-2020?
* How does the trend in birth rates vary with demographic factors such as age, race, education and marital status?
* How do the birth rates vary for different cohorts of mothers by their birth year?

While the original paper used Stata [@citeStata] for data processing and analysis in its replication package, we use R [@citeR] for all data wrangling and analysis and R packages tidyverse [@tidy], ggthemes [@ggthemes], ggprism [@ggprism] and patchwork [@patchwork] to produce the figures, kableExtra [@kableExtra] to produce the tables and haven [@haven] to read the dta files.

# Data

## Data Source and Methodology

The data on birth rates per 1,000 women ages 15-44 across all races and demographic population subgroups are obtained from the National Vital Statistics Reports for the years 2015, 2019 and 2020 [@2015birth; @2019birth; @2020birth], which they collected from birth certificates registered in all  US states and the District of Colombia. Aggregated data on the number of births for 6 different 5-year cohorts of mothers by their age and birth year is provided by @og, calculated using public birth microdata across the period 1980-1989 from the NBER Natality Database and NCHS microdata from the period 1990-2019 [@nchsR; @nber]. The birth cohorts span the years 1968-1997. The NCHS microdata is restricted use and requires an application to be obtained. We simply used the data that was supplied in the replication packageby @og.

Single-age population counts, among all races from 1969-2019 and by race and Hispanic origin from 1990-2019, is obtained from the CDC SEER database [@nci].

## Attributes

The birth data consists of variables for the year, and birth rates for each population subgroup. For example, for race there are 3 variables: birth rate for whites, birth rate for blacks and birth rate for hispanics. There is birth rate data for 4 population subgroups - age (5-year groupings), race and ethnicity, marital status and mother's level of education. We show a glimpse of the data with variables for race and marital status in table \@ref(tab:birthdata).

>>>>

---
title: "The Impact of Past Behavior Normality on Regret: A Replication Study"
subtitle: "Reproduction of 'The Puzzle of Falling US Birth Rates since the Great Recession' (Kearney, Levine & Pardue, 2022)"
author: Heng Ma & Bolin Shen
thanks: "Code and data are available at: https://github.com/Brian031205/paper_2.git. A replication of various aspects in this paper are available at: ______________    "
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Norm theory (Kahneman & Miller 1986) described a tendency for people to associate stronger regret with a negative outcome when it is a result of an exception (abnormal behavior) compared to when it is a result of routine (normal behavior). In this paper, we conducted a replication of one classic experiments on past behavior exception/routine contrasts. We successfully replicated Kahneman and Miller’s (1986) experiments with the classic hitchhiker-scenario (Part 1) and car accident-scenario (Part 2). In both cases, participants examined negative outcomes and tended to indicate a protagonist who deviated from own past behavior as more regretful than another who followed routine. In robbery scenario experiments using compensatory measures (Part 3), there is no significant difference between special and regular situations in the compensation comparison of robbery victims. However, it is showed that robberies in exceptional circumstances are considered more regrettable than robberies in regular circumstances. We discuss implications for current and future research."
output:
  bookdown::pdf_document2
toc: TRUE
bibliography: references.bib
---

@@

Regret is a pervasive and intense negative emotion, plays a pivotal role in shaping individuals' daily experiences. The recognition that other decisions or circumstances could have led to more favorable outcomes triggers regret (Byrne, 2016; Roese & Epstude, 2017; Kahneman & Tversky, 1982). Influential work by Kahneman and Miller (1986) in norm theory has highlighted the connection between abnormal behavior, heightened counterfactual thinking, and intensified feelings of regret. Their discussion of the concept of normality included experimental research illustrating various factors influencing the perception and interpretation of normality (Feldman & Albarracín, 2017; Feldman, 2018). This approach often employed past behavior as a reference point to assess the normality of an individual's actions.

Illustrating the effect of past behavior normality, Kahneman and Miller (1986) presented a thought experiment involving a car accident. Mr. White, accustomed to driving home along the same route daily, deviated from his routine one day and found himself in an accident. 82% of participants believed that Mr. White would experience more regret than those involved in similar accidents on his usual route. This bias referenced as the "exceptionality effect," serves as a basis for this study, emphasizing the regularity of past behavior.

Kahneman and Miller's (1986) norm theory provides a psychological classification of events as normal or abnormal, assuming that abnormal events elicit stronger reactions because of the increased accessibility of normal events in memory. This high availability of psychologically simulated alternatives intensifies regret over negative outcomes. Everyday behaviors, more easily recalled and psychologically emphasized, enable individuals to vividly imagine counterfactual scenarios in exceptional circumstances, thus amplifying the regret associated with deviations from past behavior (Kahneman & Miller 1986).

This study aims to directly replicate three experiments that demonstrated the exceptionality effect, representing early investigations into the influence of past behavior normality on regret. Specifically, we investigate hitchhiking and car accident scenarios from Kahneman and Miller (1986) in the first two part and replicate Miller and McFarland's (1986) Experiment 1, featuring a robbery scenario, in the part three to test norm theory. Although there is no documented direct replication of these experiments, the existing citation impact of Kahneman and Miller's work emphasizes the significance of a reexamination and detailed explanation of these effects. The study's findings are expected to contribute to a better understanding of the general nature of the exceptionality effect and the intricate influence of past behavior normality on regret.

we use R [@citeR] for all data wrangling and analysis and R packages tidyverse [@tidy], ggthemes [@ggthemes], ggprism [@ggprism] and patchwork [@patchwork] to produce the figures, kableExtra [@kableExtra] to produce the tables.

# Data
The data we used in this paper are obtained from the paper titled “The impact of past behavior normality on regret: replication and extension of three experiments of the exceptionality effect” (Kutscher & Feldman 2022). We first cleaned the data, and then analyzed them by the open-source statistical programming language R (R Core Team 2022). We used functionalities from tidyverse (Wickham et al. 2019), ggplot2 (Wickham 2016), dplyr (Wickham et al. 2022), readr (Wickham, Hester, and Bryan 2022), tibble (Müller and Wickham 2022), janitor (Firke 2021) and knitr (Xie 2014).

## Proportion for perceived regret, injunctive social norms, descriptive social norms, and negative affect
The first figure presents an overview of the perceived regret, injunctive social norms, descriptive social norms, and negative affect proportions associated with the behaviors of Mr. Smith (routine) and Mr. Jones (exception). In terms of regret, the figure highlights an obvious contrast, with Mr. Smith at 7.9% and Mr. Jones at a much higher 92.1%. A similar pattern emerges for injunctive social norms, where Mr. Smith scores 4.7%, contrasting sharply with Mr. Jones at 95.3%. Descriptive social norms show an obvious preference for routine behavior, as reflected in Mr. Smith's 90.6%, while Mr. Jones lags behind at 9.4%. Negative affect follows a parallel trend, with Mr. Smith at 7.3% and Mr. Jones at 92.7%. This figure vividly illustrates the remarkable differences in perceptions between routine and exceptional behaviors, emphasizing the notable impact on regret and social norm evaluations.

## Proportion for perceived regret and luck
Figure 2, titled "Proportions for Perceived Regret and Luck," provides a clear visual representation of the contrasts in perceived regret and luck between the behaviors of Mr. Adams (routine) and Mr. White (exception). The data shows a large difference in regret percentages, with Mr. Adams scoring at 19%, while Mr. White gets a significantly higher 81%. The figure also reveals the aspect of luck, where Mr. Adams has a proportion of 33.3%, distinctly lower than Mr. White, who gets a larger share at 66.7%. This figure effectively demonstrates the difference in perceptions regarding regret and luck associated with routine and exceptional behaviors, revealing the dynamic relationship between these two factors in the participants' evaluations of the scenarios described.

## Victim compensation and regret (with exception conditions combined)
The last figure titled "Victim Compensation and Regret (with Exception Conditions Combined)," presents an overview of two experiments, each with routine and exception conditions. The top two plots correspond to Experiment 1, while the bottom two plots represent Experiment 2, with error bars indicating standard errors. The scales range from 1 to 5 for compensation and 1 to 10 for regret. In Experiment 1, the routine condition includes 114 participants, with standard deviations of 2.37 for compensation and 1.12 for regret. The self-produced exception condition involves 115 participants, with standard deviations of 2.46 for compensation and 0.80 for regret. For the other-produced exception condition (113 participants), standard deviations are 2.76 for compensation and 0.94 for regret. Combining all exception conditions (228 participants) gets standard deviations of 2.61 for compensation and 0.87 for regret. In Experiment 2, routine conditions involve 116 participants, with standard deviations of 3.40 for compensation and 1.19 for regret. The self-produced exception condition (110 participants) shows standard deviations of 3.50 for compensation and 0.67 for regret. The other-produced exception condition (116 participants) has standard deviations of 3.66 for compensation and 0.98 for regret. Combining all exception conditions (226 participants) results in standard deviations of 3.57 for compensation and 0.86 for regret. These detailed data provide an understanding of victim compensation and regret across various experimental conditions.
